---
title: "Assignment 2: Coding Basics"
author: "Nadia Barbo"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
#1. 
#creating sequence of numbers from 1 to 100 increasing by a factor of 4
seq100 <- seq(1,100,4) 

#2. 
mean(seq100) #finding the mean of seq100
median(seq100) #finding the median of seq100

#3. 
#determining if the mean of seq100 is greater than the median of seq100 (TRUE or FALSE)
mean(seq100) > median(seq100) 
```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
#5.
studentnames <- c("Nadia", "Karina", "Abby", "Emma") #creating vector of names of students
testscores <- c(12, 87, 53, 100) #creating vector of test scores
passedtest <- c(FALSE, TRUE, TRUE, TRUE) #creating vector of students passing test or not

#6.
typeof(studentnames) #studentnames is a character (chr) vector
typeof(testscores) #testscores is a double/numeric (num) vector
typeof(passedtest)#passed test is a logical (logi) vector

#7. and 8.
#combining previously created vectors into data frame and naming columns of the data frame
testresults.df <- data.frame("Student.Names" = studentnames, "Test.Scores" = testscores, 
                             "Passed.Test" = passedtest) 
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: A data frame is different from a matrix because a data frame can be composed of columns/vectors that are different modes from one another (but they all need to be the same length), while in a matrix, all of the columns need to be the same mode and same length.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

```{r}
#10.
#creating function where you input a vector and if a value in that vector is >=50 it will 
#print TRUE in it's place and if it is <50 it will print FALSE
passed <- function(grade.vector){
    result <- ifelse(grade.vector>=50, TRUE, FALSE)
    print(result)
}
```

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
#11.
#running the function I just created on testscores vector
passed(testscores)
```


12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: The "ifelse" option worked because it is able to check multiple values from a vector at one time, allowing us to input a vector and print the outputs for each value in that vector. The "if" statement can only check one element in a vector at a time, so it cannot evaluate the entire vector.

