---
title: 'A07 - Crafting Reports'
author: "Nadia Barbo" 
date: "Spring 2023"
output: 
  pdf_document:
    toc: TRUE
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

## Objectives:

1.  More practice with R code chunk options
2.  Gain proficiency with figures, tables (w/`Kable`) table of contents, etc.
3.  Debugging knitting issues

## Directions

1.  Rename this file `<FirstLast>_A07_CraftingReports.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the tasks, **creating code and output** that fulfill each instruction.
4.  Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  **Be sure that you also commit and push your final Rmd document to your GitHub account**.

## Task 1 - Basic Markdown

Create a table below summarizing the metadata of the EPA Air Quality data. The first column will be the metadata attribute item name: "Source", "Date", and "Filename". And the second column will include the metadata values: "EPA Air Quality SYstem (AQS)", "2018-2019", and "EPAair_O3_PM25_NC1819_Processed.csv". The first column should be aligned to the right and the second to the left.

```{r Task1, echo=FALSE, message=FALSE, warning=FALSE}
Item_Name <- c('Source', 'Date', 'Filename') #variables for first column
Value <- c('EPA Air Quality System (AQS)', '2018-2019', 'EPAair_O3_PM25_NC1819_Processed.csv') #variables for second column
for_table.df <- data.frame(Item_Name, Value) #combining two lists into a dataframe

knitr::kable(for_table.df, #using kable function to make dataframe into table
      align = 'rl', #aligning first column to the right and second column to the left
      col.names = c('Item Name', 'Value')) #naming columns in table
```

------------------------------------------------------------------------

## Task 2 - Import packages and data, suppressing messages

Set the following R code chunk so that it runs when knit, but no messages, errors, or any output is shown. The code itself should be displayed.

```{r project.setup, message=FALSE, warning=FALSE, error=FALSE}
#Import libraries
library(tidyverse);library(lubridate);library(here);library(knitr)

#Import EPA data (from the processed_KEY folder) & fix dates
epa_data <- read.csv(
  here("data","processed_KEY","EPAair_O3_PM25_NC1819_Processed.csv"),
  stringsAsFactors = TRUE) %>% 
  mutate(Date = ymd(Date))
```

------------------------------------------------------------------------

## Task 3: Creating tables

Set the following R code chunk to display two tables, using knitr's `kable()` function, one listing the mean PM2.5 concentrations for each county, and the other the same except for Ozone. The titles should be "Mean Particulates (2.5mm)" and "Mean Ozone", respectively. And the column names should be "County" and "µg/m3" for both tables.

Customize the chunk options such that the code is run but is not displayed in the knitted document. The output, however, should be displayed.

> **TIPS:**
>
> -   Use `"$\\mu g/m^3$"` as a column name to generate a nicely formatted string via markdown/MathJax notation
>
> -   If your output table spans across two pages, try inserting a new line (via `\newpage`) in the markdown just before your code chunk.

\newpage

```{r data.summary, echo=F}
#Compute mean PM2.5 concentration for each county, sorted by value
summary_PM25 <- epa_data %>% 
  drop_na(PM2.5) %>% 
  group_by(COUNTY) %>% 
  summarize(mean_PM2.5 = mean(PM2.5)) %>% 
  arrange(mean_PM2.5)

#Compute mean ozone concentration for each county, sorted by value
summary_Ozone <- epa_data %>% 
  drop_na(Ozone) %>% 
  group_by(COUNTY) %>% 
  summarize(
    mean_Ozone = mean(Ozone)) %>% 
  arrange(mean_Ozone)

#Use the `kable` function to produce the 2 tables
kable(summary_PM25, #using kable to make table of PM2.5 summary data
      caption = "Mean Particulates (2.5mm)", #adding title
      col.names = c("County", "$\\mu g/m^3$")) #adding column names

kable(summary_Ozone, #using kable to make table of Ozone summary data
      caption = "Mean Ozone", #adding title
      col.names = c("County", "$\\mu g/m^3$")) #adding column names
```

\newpage

## Task 3: Plots

Create two separate code chunks that create boxplots of the distribution of Ozone levels by month using, one for only records collected in 2018 and one for records in 2019. Customize the chunk options such that the final figures are displayed but not the code used to generate the figures. In addition, the plots aligned on the left side of the page and set the figure heights so both plots fit on the same page with minimal space remaining. Lastly, add a `fig.cap` chunk option to add a caption (title) to your plot that will display underneath the figure.

```{r Task3 Plot1, fig.align='left', fig.height = 2.85, fig.cap = "Monthly Ozone, 2018", echo=FALSE, warning=FALSE}
ggplot(epa_data %>% filter(year(Date) == "2018"), #filtering for only data in 2018
       aes(
         x=factor(month(Date), levels = 1:12, labels = month.abb), #assigning x variable (as a factor) and adding month abbs
         y=Ozone)) + #assigning y variable
  geom_boxplot() + #making a boxplot
  labs(x="Month", y=expression(paste('Concentration (',mu,'g/m'^3,')')), title="Recorded Ozone by Month (2018)") #adding axis labels and title
```

```{r Task3 Plot2, fig.align='left', fig.height = 2.85, fig.cap = "Monthly Ozone, 2019", echo=FALSE, warning=FALSE}
ggplot(epa_data %>% filter(year(Date) == "2019"), #filtering for only data in 2019
       aes(
         x=factor(month(Date), levels = 1:12, labels = month.abb), #assigning x variable (as a factor) and adding month abbs
         y=Ozone)) + #assigning y variable
  geom_boxplot() + #making a boxplot
  labs(x="Month", y=expression(paste('Concentration (',mu,'g/m'^3,')')), title="Recorded Ozone by Month (2019)") #adding axis labels and title
```

\newpage

## Task 4: Knit and submit.

Add a table of contents to your document and knit to a PDF. Submit your PDF to Sakai, but also be sure to commit and push your Rmd file used to create this knit document to GitHub. In the section below, add a link to your GitHub repository.

### Git Repository
https://github.com/nadiabarbo/EDA-Spring2023
