---
title: 'Assignment 9: Spatial Analysis in R'
author: "Nadia Barbo"
output:
  html_document:
    df_print: paged
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics (ENV872L) on spatial analysis.

## Directions
1. Rename this file `<FirstLast>_A09_SpatialAnalysis.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Use the lesson as a guide. It contains code that can be modified to complete the assignment.
4. Work through the steps, **creating code and output** that fulfill each instruction.
5. Be sure to **answer the questions** in this assignment document. Space for your answers is provided in this document and is indicated by the ">" character. If you need a second paragraph be sure to start the first line with ">". You should notice that the answer is highlighted in green by RStudio.
5. When you have completed the assignment, **Knit** the text and code into a single HTML file.


## DATA WRANGLING

### Set up your session

1. Import libraries: tidyverse, sf, leaflet, and mapview
2. Check your working directory

```{r Prepare workspace, warning=F, error=F, message=F}
#1. 
#importing libraries
library(tidyverse)
library(lubridate)
library(sf)
library(mapview); mapviewOptions(fgb = FALSE)

#2. 
#checking working directory
getwd()

#also setting my theme 
mytheme <- theme_bw() + #building my theme and making base font size 14
  theme(plot.title = element_text(hjust = 0.5), #making plot title centered
        plot.subtitle = element_text(hjust = 0.5), #making plot subtitle centered
        panel.background = element_rect(fill = 'white'), #making panel background off white
        plot.background = element_rect(fill = "white")) #making plot background off white
theme_set(mytheme) #setting my theme to be default
```

### Read (and filter) county features into an sf dataframe and plot

In this exercise, we will be exploring stream gage height data in Nebraska corresponding to floods occurring there in 2019. First, we will import from the US Counties shapefile we've used in lab lessons, filtering it this time for just Nebraska counties. Nebraska's state FIPS code is `31` (as North Carolina's was `37`).

3. Read the `cb_2018_us_county_20m.shp` shapefile into an sf dataframe, filtering records for Nebraska counties (State FIPS = 31)
4. Reveal the dataset's coordinate reference system
5. Plot the records as a map (using `mapview` or `ggplot`)

```{r Read the county data into an sf dataframe, warning=F, error=F, message=F}
#3. Read in Counties shapefile into an sf dataframe, filtering for just NE counties
NE_counties_sf <- st_read('./Data/Spatial/cb_2018_us_county_20m.shp') %>% #load in the counties shapefile
  filter(STATEFP == 31) #filter for just Nebraska counties

#4. Reveal the CRS of the counties features
st_crs(NE_counties_sf) #showing coordinate reference system

#5. Plot the data
ggplot(data=NE_counties_sf) + #plot NE counties
  geom_sf(fill="lightblue", color="black") + #as a map
  labs(title="Nebraska Counties") #adding a title

###NOTE: Whenever I make a map with ggplot the degrees symbols appear as boxes - I talked with John about this in the Slack and he said that it seems to be a font issue and that it is okay to submit it
```

6. What is the EPSG code of the Counties dataset? Is this a geographic or a projected coordinate reference system? (Or, does this CRS use angular or planar coordinate units?) To what datum is this CRS associated? (Tip: look the EPSG code on <https://spatialreference.org>) 

> ANSWER: 
> The EPSG code for the Counties dataset is 4269. This is a projected coordinate reference system. This CRS is associated with the North American Datum 1983 (NAD83) datum.


### Read in gage locations csv as a dataframe, then display the column names it contains

Next we'll read in some USGS/NWIS gage location data added to the `Data/Raw` folder. These are in the `NWIS_SiteInfo_NE_RAW.csv` file.(See `NWIS_SiteInfo_NE_RAW.README.txt` for more info on this dataset.)

7. Read the NWIS_SiteInfo_NE_RAW.csv file into a standard dataframe.

8. Display the column names of this dataset.

```{r Read in USGS Gage Locations, warning=F, error=F, message=F}
#7. Read in gage locations csv as a dataframe
gage_locations_df <- read.csv('./Data/Raw/NWIS_SiteInfo_NE_RAW.csv', #reading in dataframe as CSV
  colClasses = c('site_no' = 'factor')) #reading site_no column as factor

#8. Reveal the names of the columns
colnames(gage_locations_df) #showing column names of gage dataframe
```

9. What columns in the dataset contain the x and y coordinate values, respectively?  

> ANSWER:
> The dec_long_va contains the x coordinates and the dec_lat_va contains the y coordinates.


### Convert the dataframe to a spatial features ("sf") dataframe

10. Convert the dataframe to an sf dataframe.
 * Note: These data use the same coordinate reference system as the counties dataset

11. Display the column names of the resulting sf dataframe

```{r Convert dataframe to sf dataframe, warning=F, error=F, message=F}
#10. Convert to an sf object
gage_locations_sf <- gage_locations_df %>% 
  st_as_sf(coords = c('dec_long_va','dec_lat_va'), crs=4269) #converting df to sf and defining coordinates

#11. Re-examine the column names
colnames(gage_locations_sf) #checking column names
```

12. What new field(s) appear in the sf dataframe created? What field(s), if any, disappeared?

>ANSWER: 
> The "geometry" field is a new field that appears at the end of the sf dataframe, defining the spatial coordinates we set when converting the dataframe to an sf object. The "dec_lat_va" and "dec_long_va" columns disappear in the sf object. 


### Plot the gage locations on top of the counties

13. Use `ggplot` to plot the county and gage location datasets.
 * Be sure the datasets are displayed in different colors
 * Title your plot "NWIS Gage Locations in Nebraska"
 * Subtitle your plot with your name

```{r Plot the spatial features, warning=F, error=F, message=F}
#13. Plot the gage locations atop the county features
ggplot() + #make plot
  geom_sf(data=NE_counties_sf, fill="lightblue", color="black") + #as a NE counties map
  geom_sf(data=gage_locations_sf, color="darkblue") + #with the gage locations map too
  labs(title="NWIS Gage Locations in Nebraska", subtitle = "Nadia Barbo") #adding title and subtitle
```


### Read in the gage height data and join the site location data to it.

Lastly, we want to attach some gage height data to our site locations. I've constructed a csv file listing many of the Nebraska gage sites, by station name and site number along with stream gage heights (in meters) recorded during the recent flood event. This file is titled `NWIS_SiteFlowData_NE_RAW.csv` and is found in the Data/Raw folder.

14. Read in and wrangle the `NWIS_SiteFlowData_NE_RAW.csv` dataset: 
  * Be sure the `site_no` column is read in as a factor
  * Group by `site_no` and `station_nm`, computing the mean `gage_ht` across groups values
15. Show the column names .
16. Join our site information (already imported above) to these gage height data.
 * The `site_no` and `station_nm` can both/either serve as joining attributes.
 * Construct this join so that the result only includes spatial features where both tables have data.
17. Show the column names in this resulting spatial features object
18. Show the dimensions of the resulting joined dataframe

```{r Joining gage height to gage location, warning=F, error=F, message=F}
#14. Read the site flow data into a data frame
NE_flowdata_df <- read.csv('./Data/Raw/NWIS_SiteFlowData_NE_RAW.csv', #reading in dataframe as CSV
  colClasses = c('site_no' = 'factor')) %>% #reading site_no column as factor
  group_by(site_no, station_nm) %>% #grouping data
  summarize(mean_gage_ht = mean(gage_ht)) #adding mean gage height variable

#15. Show the column names
colnames(NE_flowdata_df) #showing column names

#16. Join the flow data to our NWIS gage location spatial dataframe
gages_flowandlocation <- gage_locations_sf %>%  #making new dataframe
  left_join(NE_flowdata_df, by = c("site_no", "station_nm")) %>% #joining location and flow data
  na.omit() #removing rows with NAs (Can we use drop_na?)

#17. Show the column names of the joined dataset
colnames(gages_flowandlocation) #showing column names

#18. Show the dimensions of this joined dataset
dim(gages_flowandlocation) #showing dimensions
```

### Map the pattern of gage height data

Now we can examine where the flooding appears most acute by visualizing gage heights spatially.

19. Plot the gage sites on top of counties (using `mapview`, `ggplot`, or `leaflet`) 
 * Show the magnitude of gage height by color, shape, other visualization technique.

```{r Mapping pattern of gage height data, warning=F, error=F, message=F}
#Map the points, colored by gage height
mapview(NE_counties_sf, col.regions="lightpink", layer.name="NE Counties") + #making map of NE counties
  mapview(gages_flowandlocation, legend=TRUE, alpha.regions=1, layer.name="Mean Gage Height (meters)", zcol="mean_gage_ht", col.regions=(mapviewPalette("mapviewSpectralColors"))(114), cex=4) #making map of gage locations and coloring them by gage height

#Map the points, sized by gage height - but I couldn't get the legend to format how I wanted it.
# mapview(NE_counties_sf, col.regions="lightpink", alpha.region=0.75, layer.name="NE Counties") +
#   mapview(gages_flowandlocation, cex=gages_flowandlocation$mean_gage_ht, alpha.regions=0.75, layer.name="Mean Gage Height (meters)", col.regions="yellow") #how do i add the legend showing point sizes?
```

---

## SPATIAL ANALYSIS

Up next we will do some spatial analysis with our data. To prepare for this, we should transform our data into a projected coordinate system. We'll choose UTM Zone 14N (EPGS = 32614).


### Transform the counties and gage site datasets to UTM Zone 14N

20. Transform the counties and gage sf datasets to UTM Zone 14N (EPGS = 32614).
21. Using `mapview` or `ggplot`, plot the data so that each can be seen as different colors

```{r Transforming counties and gage sites to UTM Zone 14, warning=F, error=F, message=F}
#20 Transform the counties and gage location datasets to UTM Zone 14
NE_counties_sf_utm <- st_transform(NE_counties_sf, crs = 32614) #transforming NE counties to UTM Zone 14
gage_locations_sf_utm <- st_transform(gage_locations_sf, crs = 32614) #transforming gage locations to UTM Zone 14

#21 Plot the data
ggplot() + #make plot
  geom_sf(data=NE_counties_sf_utm, fill="lightblue", color="black") + #as a NE counties map
  geom_sf(data=gage_locations_sf_utm, color="darkblue") + #with the gage locations map too
  labs(title="NWIS Gage Locations in Nebraska", subtitle = "Nadia Barbo") #adding title and subtitle
```


### Select the gages falling within a given county

Now let's zoom into a particular county and examine the gages located there. 

22. Select Douglas county from your projected county sf dataframe 

23. Select the gage sites falling `within` that county
 * Use either matrix subsetting or tidy filtering 

24. Create a plot showing:
 * all Nebraska counties,
 * the selected county, 
 * and the gage sites in that county

```{r Filtering for Douglas County, warning=F, error=F, message=F}
#22 Select the county
NE_DouglasCounty_sf_utm <- NE_counties_sf_utm %>%
  filter(NAME == "Douglas") #filtering for just Douglas County in the NE_counties_sf dataframe

#23 Select gages within the selected county
douglas_gages <- gage_locations_sf_utm %>% 
  st_filter(NE_DouglasCounty_sf_utm, .pred = st_intersects)

#24 Plot
#using mapview
mapview(NE_counties_sf, col.regions = "lightpink", layer.name = "NE Counties", alpha.regions = 0.75) + #making map of NE counties
  mapview(NE_DouglasCounty_sf_utm, col.regions = "yellow", layer.name = "Douglas County, NE", alpha.regions = 0.75) + #map of just Douglas County
  mapview(douglas_gages, alpha.regions = 1, layer.name = "Douglas County Gages", col.regions = "red", cex = 3) #map of gages in Douglas County
```
