---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Nadia Barbo"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r setup, include=FALSE} 
#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1

#checking working directory
getwd()

#loading necessary packages
library(tidyverse); library(lubridate); library(here); library(agricolae); library(ggsci)

#loading in NTL-LTER raw data files for chemistry/physics
Lake_Nutrients <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)

#changing date columns to date format
Lake_Nutrients$sampledate <- mdy(Lake_Nutrients$sampledate)

#2
nadia_theme <- theme_classic() + 
  theme(
    axis.title = element_text(color = "pink3", #making axis titles pink
                              family = "sans",#making axis titles 'sans' font
                              face = 'bold', #making axis titles bold
                              size = 14), #making axis titles larger (size 14)
    plot.title = element_text(color = "pink3", #making plot title pink
                              family = "sans", #making plot title 'sans' font
                              hjust = 0.5, #making plot title centered
                              face = 'bold', #making plot title bold
                              size = 16), #making plot title size 16
    
    panel.background = element_rect(fill = 'whitesmoke'), #making panel background off white
    plot.background = element_rect(fill = "whitesmoke", #making plot background off white
                                   color = "pink2"), #making plot background outline pink
    axis.line.x.bottom = element_line(linewidth = 1, #making bottom axis line thinner
                                      color='pink2'), #making bottom axis pink
    axis.line.y.left = element_line(linewidth = 1, #making left axis line thinner
                                    color='pink2'), #making left axis pink
    axis.ticks = element_line(color = 'pink3', #making axis tick marks pink
                              linewidth = 1), #making axis tick marks size 1
    axis.text = element_text(color='pink3', #making axis labels pink
                             face = 'bold'), #making axis labels bold
    
    panel.grid.major = element_line(linetype = 2, #adding grid lines and making them dashed
                                    linewidth = 1, #making grid lines size 1
                                    color="pink2"), #making grid lines pink

    legend.background = element_rect(fill = 'whitesmoke', #making the legend background off white
                                     color='pink2'), #making the legend outline pink
    legend.text = element_text(family = 'sans', #making the legend text 'sans' font 
                               size = 12, #making the legend text size 12
                               color = "pink3", #making the legend text pink
                               face='bold'), #making the legend text bold
    legend.title = element_text(color="pink3", #making the legend title pink
                                face="bold", #making the legend title bold
                                size=14, #making the legend title size 14
                                family ='sans'), #making the legend title 'sans' font
    legend.position = 'bottom' #putting the legend on the bottom of the plot
    )

theme_set(nadia_theme) #setting my theme as the default theme
```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:

> Answer:
H0: mu_1 = mu_2 = mu_n (The mean lake temperature during July does not change with depth across all lakes)
Ha: mu_1 != mu_2 != mu_n (The mean lake temperature during July does change with depth across all lakes)


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot, fig.height = 6, fig.width = 10}
#4
Lake_Nutrients_processed <-  Lake_Nutrients %>% #creating processed data frame
  filter(month(sampledate) == "7") %>% #filtering for only samples in July
  select(lakename, year4, daynum, depth, temperature_C) %>% #selecting only the columns we want
  drop_na() #dropping NAs

#5
ggplot(Lake_Nutrients_processed, aes(x=depth, y=temperature_C)) + #setting variables to plot
  geom_point(color='palegreen3') + #make scatterplot with green points
  geom_smooth(method="lm", color='black', se=FALSE) + #adding best fit line
  ylim(0,35) + #limiting temperature values to 0-35
  labs(x="Depth (m)", #adding x label
       y="Temperature (C)", #adding y label
       title="Temperature by Depth Changes in July Months") #adding title
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure suggests that as the depth increases, the temperature decreases. This does not look like a linear relatioship - it looks more like an exponential decrease in temperature as depth increases.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
Temperature.regression <- lm(Lake_Nutrients_processed$temperature_C ~ Lake_Nutrients_processed$depth) #performing linear regression
summary(Temperature.regression) #show summary of linear regression

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The relationship between decreasing temperatures and increasing depths was found to be statistically significant (p value < 2e-16). About 73.87% of the variability in temperature is explained by changes in depth (given by the R-squared value). These findings are based on 9726 degrees of freedom. According to the linear regression, temperature is predicted to decrease by 1.95 degrees Celcius for every 1m change in depth.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
Temperature_AIC <- lm(data = Lake_Nutrients_processed, temperature_C ~ year4 + daynum + 
              depth) #running linear multiple regression using 3 designated explanatory variables 
step(Temperature_AIC) #using step function to determine which variables are best suited to predict temperature

#10
Temperature_model <- lm(data = Lake_Nutrients_processed, temperature_C ~ year4 + daynum + 
              depth) #the step function kept all four variables
summary(Temperature_model) #summary of best fit model

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The step function/AIC method suggested we keep all three explanatory variables in the multiple regression model to predict temperature: year4, daynum, and depth. This multiple regression model explains 74.11% of the variance, which is slightly more than the linear regression model using only depth as an explanatory variable (73.87%). 



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
Temp_Lake.anova <- aov(data = Lake_Nutrients_processed, temperature_C ~ lakename) #running ANOVA model
summary(Temp_Lake.anova) #showing summary of anova model

Temp_Lake.anova2 <- lm(data = Lake_Nutrients_processed, temperature_C ~ lakename) #running linear regression model
summary(Temp_Lake.anova2) #showing summary of linear regression model
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: There is a significant different in mean temperature among the lakes, with a p value < 2e-16. When looking at the significance of individual lakes on temperature, all lakes were highly significant (p values labeled as either *** or **). According to the linear model, 4.87% of temperature variance can be explained by the lakename variable. The linear model also suggests that the following lakes decrease temperature by the corresponding amount for every 1m change in depth: Crampton Lake - decrease of 2.31 degrees, East Long Lake - decrease of 7.40 degrees, Hummingbird Lake - decrease of 6.89 degrees, Paul Lake - decrease of 3.85 degrees, Peter Lake - decrease of 4.35 degrees, Tuesday Lake - decrease of 6.60 degrees, Ward Lake - decrease of 3.21 degrees, and West Long Lake - decrease of 6.09 degrees. 


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2, fig.height = 6, fig.width = 10}
#14.
ggplot(Lake_Nutrients_processed, aes(x=depth, y=temperature_C, color=lakename)) + #setting variables to plot (colored by lakename)
  geom_point() + #making scatterplot
  scale_color_npg() + #setting color of points
  geom_smooth(method="lm", color='black', se=FALSE) + #adding line of best fit
  ylim(0,35) + #setting y axis limits
  labs(x="Depth (m)", #naming x axis
       y="Temperature (C)", #naming y axis
       title="Depth and Temperature Changes in July Months per Lake", #adding title
       color="Lake Name") #changing legend title

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(Temp_Lake.anova) #running Tukey HSD test

Temps.groups <- HSD.test(Temp_Lake.anova, "lakename", group = TRUE) #finding groupings from Tukey HSD test
Temps.groups #showing those groupings
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Paul Lake and Ward Lake have statistically the same mean temperature as Peter Lake. There are no lakes that have a mean temperature that is statistically distict from all other lakes. 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: If we were just looking at Peter Lake and Paul Lake, we could also run a two sample t-test to determine if the two lakes have distinct mean temperatures. A two sample t-test uses a continuous response variable (temperature) and one categorical explanatory variable with two categories (lakename - in this case the two categories are Peter Lake and Paul Lake). This two sample t-test will test the mean response variable of the two categories and determine if they are statistically the same or different.



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
#wranging data
Lake_Nutrients_July_CramptonWard <- Lake_Nutrients_processed %>% #creating processed data frame
  filter(lakename %in% c("Crampton Lake", "Ward Lake")) #filtering for only samples from Crampton Lake and Ward Lake

Temp.twosample <- t.test(Lake_Nutrients_July_CramptonWard$temperature_C ~ Lake_Nutrients_July_CramptonWard$lakename) #running two-sample t-test
Temp.twosample #showing results of t-test
```

>Answer: The t-test tells us that we cannot reject the null hypothesis that the mean temperatures of Crampton Lake and Ward Lake are the same. In other words, the mean temperatures within these two lakes are not statistically different from one another (they are statistically equal). We cannot reject the null hypothesis because the p-value from the t-test is 0.2649. This does match the answer for part 16 because Crampton Lake and Ward Lake were both sorted into group b, meaning that their mean temperatures were statistically the same as one another. You can also confirm this because the estimated mean temperatures of the two lakes (given by the t-test) are within the same within the 95% confidence interval. 
