---
title: "Assignment 4: Data Wrangling"
author: "Nadia Barbo"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
#install.packages('Rtools')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```


## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions

1. Rename this file `<FirstLast>_A04_DataWrangling.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Feb 20th @ 5:00pm.

## Set up your session 

1a. Load the `tidyverse`,  `lubridate`, and `here` packages into your session.

1b. Check your working directory.

1c. Read in all four raw data files associated with the EPA Air dataset, being sure to set string columns to be read in a factors. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Apply the `glimpse()` function to reveal the dimensions, column names, and structure of each dataset. 

```{r Questions 1-2, message = FALSE}
#1a 
library(tidyverse); library(lubridate); library(here) #loading necessary packages

#1b 
getwd() #checking my working directory

#1c 
NC_O3_2018 <- read.csv("./Data/Raw/EPAair_O3_NC2018_raw.csv", stringsAsFactors = TRUE) #loading in EPA O3 NC 2018 dataset
NC_O3_2019 <- read.csv("./Data/Raw/EPAair_O3_NC2019_raw.csv", stringsAsFactors = TRUE) #loading in EPA O3 NC 2019 dataset
NC_PM25_2018 <- read.csv("./Data/Raw/EPAair_PM25_NC2018_raw.csv", stringsAsFactors = TRUE) #loading in EPA PM2.5 NC 2018 dataset
NC_PM25_2019 <- read.csv("./Data/Raw/EPAair_PM25_NC2019_raw.csv", stringsAsFactors = TRUE) #loading in EPA PM2.5 NC 2019 dataset

#2 
glimpse(NC_O3_2018) #revealing dimensions, column names, and structure of EPA O3 NC 2018 dataset
glimpse(NC_O3_2019) #revealing dimensions, column names, and structure of EPA O3 NC 2019 dataset
glimpse(NC_PM25_2018) #revealing dimensions, column names, and structure of EPA PM2.5 NC 2018 dataset
glimpse(NC_PM25_2019) #revealing dimensions, column names, and structure of EPA PM2.5 NC 2019 dataset
```


## Wrangle individual datasets to create processed files.

3. Change date columns to be date objects.

4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE

5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).

6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r Questions 3-6}
#3
NC_O3_2018$Date <- mdy(NC_O3_2018$Date) #Changing date column in EPA O3 NC 2018 dataset (which are originally factors) to be date objects
NC_O3_2019$Date <- mdy(NC_O3_2019$Date) #Changing date column in EPA O3 NC 2019 dataset (which are originally factors) to be date objects
NC_PM25_2018$Date <- mdy(NC_PM25_2018$Date) #Changing date column in EPA PM2.5 NC 2018 dataset (which are originally factors) to be date objects
NC_PM25_2019$Date <- mdy(NC_PM25_2019$Date) #Changing date column in EPA PM2.5 NC 2019 dataset (which are originally factors) to be date objects

#4
NC_O3_2018_W <- 
  NC_O3_2018[, c("Date", "DAILY_AQI_VALUE", "Site.Name", "AQS_PARAMETER_DESC", "COUNTY", "SITE_LATITUDE", "SITE_LONGITUDE")] #Selecting for specific columns in EPA O3 NC 2018 dataset
NC_O3_2019_W <- 
  NC_O3_2019[, c("Date", "DAILY_AQI_VALUE", "Site.Name", "AQS_PARAMETER_DESC", "COUNTY", "SITE_LATITUDE", "SITE_LONGITUDE")] #Selecting for specific columns in EPA O3 NC 2019 dataset
NC_PM25_2018_W <- 
  NC_PM25_2018[, c("Date", "DAILY_AQI_VALUE", "Site.Name", "AQS_PARAMETER_DESC", "COUNTY", "SITE_LATITUDE", "SITE_LONGITUDE")] #Selecting for specific columns in EPA PM2.5 NC 2018 dataset
NC_PM25_2019_W <- 
  NC_PM25_2019[, c("Date", "DAILY_AQI_VALUE", "Site.Name", "AQS_PARAMETER_DESC", "COUNTY", "SITE_LATITUDE", "SITE_LONGITUDE")] #Selecting for specific columns in EPA PM2.5 NC 2019 dataset

#5
NC_PM25_2018_W$AQS_PARAMETER_DESC <- "PM2.5" #filling all cells in AQS_PARAMETER_DESC with "PM2.5" in the EPA PM2.5 NC 2018 dataset
NC_PM25_2019_W$AQS_PARAMETER_DESC <- "PM2.5" #filling all cells in AQS_PARAMETER_DESC with "PM2.5" in the EPA PM2.5 NC 2019 dataset

#6
write.csv(NC_O3_2018_W, row.names = FALSE, file = "./Data/Processed/EPAair_O3_NC2018_processed.csv") #saving processed EPA O3 NC 2018 dataset in Processed folder
write.csv(NC_O3_2019_W, row.names = FALSE, file = "./Data/Processed/EPAair_O3_NC2019_processed.csv") #saving processed EPA O3 NC 2019 dataset in Processed folder
write.csv(NC_PM25_2018_W, row.names = FALSE, file = "./Data/Processed/EPAair_PM25_NC2018_processed.csv") #saving processed EPA PM2.5 NC 2018 dataset in Processed folder
write.csv(NC_PM25_2019_W, row.names = FALSE, file = "./Data/Processed/EPAair_PM25_NC2019_processed.csv") #saving processed EPA PM2.5 NC 2019 dataset in Processed folder
```


## Combine datasets

7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code.

8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

  * Include all sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` can figure out common factor levels - but it will include sites with missing site information...) 

  * Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site name, AQS parameter, and county. Take the mean of the AQI value, latitude, and longitude.
  
  * Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
  
  * Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.

10. Call up the dimensions of your new tidy dataset.

11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1819_Processed.csv"

```{r Questions 7-11}
#7 
EPAair_O3_PM25_NC1819 <- rbind(NC_O3_2018_W, NC_O3_2019_W, NC_PM25_2018_W, NC_PM25_2019_W) #Combining the four datasets

#8 
EPAair_O3_PM25_NC1819_W <- 
  EPAair_O3_PM25_NC1819 %>%
  #filtering by site names that all four data frames have in common
  filter(Site.Name %in% c("Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School")) %>%
  #using the split-apply-combine strategy (split using group_by)
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY) %>%
  ##using the split-apply-combine strategy - using summarise to add mean values to data frame
  summarise(MEAN_AQI_VALUE = mean(DAILY_AQI_VALUE), 
            MEAN_LATITUDE = mean(SITE_LATITUDE), 
            MEAN_LONGITUDE = mean(SITE_LONGITUDE)) %>%
  #adding columns for "Month" and "Year"
  mutate(Month = month(Date), Year = year(Date))

#9
EPAair_O3_PM25_NC1819_processed <- 
  EPAair_O3_PM25_NC1819_W %>%
  pivot_wider(names_from = AQS_PARAMETER_DESC, 
              values_from = MEAN_AQI_VALUE) #widening data set so AQI values for ozone and PM2.5 are in separate columns

#10
dim(EPAair_O3_PM25_NC1819_processed) #calling up the dimensions of the data set

#11
write.csv(EPAair_O3_PM25_NC1819_processed, row.names = FALSE, file = "./Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv") #saving processed data set in Processed folder
```


## Generate summary tables

12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where mean **ozone** values are not available (use the function `drop_na` in your pipe). It's ok to have missing mean PM2.5 values in this result. 

13. Call up the dimensions of the summary dataset.

```{r Questions 12-13}

#12
EPAair_O3_PM25_NC1819_Summary <- 
  EPAair_O3_PM25_NC1819_processed %>%
  #using the split-apply-combine strategy (split using group_by)
  group_by(Site.Name, Month, Year) %>%
  ##using the split-apply-combine strategy - using summarise to add mean values to data frame
  summarise(MEAN_AQI_OZONE = mean(Ozone), 
            MEAN_AQI_PM25 = mean(PM2.5)) %>%
  #removing instances where mean ozone values are not available
  drop_na(MEAN_AQI_OZONE)


#13
dim(EPAair_O3_PM25_NC1819_Summary) #calling up the dimensions of the data set
```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: We use the 'drop_na' function instead of the 'na.omit' function because it is difficult to select a single column to remove NAs from using na.omit while drop_na can easily have a column assigned to it. Therefore, drop_na allows us to keep NAs from other columns and only drop rows that have NAs in a selected column (in this case Ozone) while using omit.na and not specifically omitting NAs from only the Ozone column would also omit NAs from all other columns within the dataframe.

